Meeting the Challenges of Modeling Astrophysical Thermonuclear Explosions: \castro, \maestro, and the \amrex\ Astrophysics Suite

### 2017-10-20 astronum_2017/astronum-amrex.tex
John's edits

The largest open theoretical question for SNe Ia is---what is the
nature of the progenitor?  About 20 years ago, the community had
mostly converged up the single-degenerate scenario---a
Chandrasekhar-mass C/O white dwarf that accretes from its companion,
eventually leading to runaway at the center that burns through the
star (see~\cite{hillebrandtniemeyer2000} for the state of the field at
that time).  Since then, the wealth of observations has show that
there is a lot of diversity in SNe Ia, and searches for progenitor
systems have not shown that Chandra-mass white dwarfs can explain most
SNe Ia.  Today, merging white dwarfs (the double degenerate scenario)
have perhaps become the most popular model.  Other progenitors, like
He burning on the surface of a sub-Chandra white dwarf have also seen
interest in explaining some of the observed diversity.
See~\cite{araa-maoz} for a review.

There are open questions in all of these scenarios which can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is, can a deflagration
transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
possible to create a detonation in the thin surface He layer?  For
double degenerates, does the burning take place promptly? and can we
avoid an accretion-induced collapse to a neutron star.  For many of
these, we also need to address numerical issues, such as whether it is
possible to accurately model the ignition of a detonation with the
spatial resolution we can attain?  These are some of the questions we
seek to answer.

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model has \amrex\ managing the grid data-structures
and parallel communications, and it calls the computational kernels on
a patch-by-patch basis.  The core library is written in \cpp\ with the
computational kernels written in Fortran\footnote{Currently
  \maestro\ is using an older pure-Fortran interface, but this will be
  ported to the \cpp\ \amrex\ this coming year.}---this allows us to take
advantage of the strengths of both languages. \amrex\ supports subcycling
in time, which we use in \castro.

\amrex\ uses a hybrid MPI + OpenMP approach to parallelism.
Distribution of grid patches to nodes using MPI provides a natural
coarse-grained approach to distributing the computational work, while
threading of loops over zones in a grid using OpenMP provides
effective fine-grained parallelization.  For \castro\, tiling is used
to logically subdivide AMR patches to increase the parallelism for
OpenMP, without incurring the memory overheads that using many more
smaller grids would bring~\cite{tiling}.  This strategy is especially
important for many core architectures like the Intel Phi.  Ongoing
development is being done to support GPU offloading, using managed
memory provided by the latest generations of GPUs.

There are many application codes built on \amrex, including those in
combustion, ... \MarginPar{domains?} In astrophysics, these include \maestro\ and
\castro\ for stellar and nuclear astrophysics applications and
\nyx~\cite{nyx} for cosmological applications.  Here we focus on the
former two.

@@

The state is advanced using a second-order accurate projection method.
Fluid quantities are advected using an unsplit Godunov method, with
reactions incorporated via operator splitting.  The provisional
velocities are then projected onto the space that satisfies the
divergence constraint.  The projections involve elliptic solves,
solving a variable-coefficient Poisson problem, computed using a
multigrid algorithm.  Low speed methods have seen a lot of recent
development~\cite{kleinpauluis,vasil:2013}, which has been
incorporated into \maestro.

@@

\castro~\cite{castro,castroII,castroIII} is a fully-compressible
radiation hydrodynamics code that supports arbitrary equations of
state, nuclear reaction networks, and Poisson gravity using geometric
multigrid.  The main hydrodynamics scheme in \castro\ is an unsplit
piecewise parabolic method.  The radiation solver in \castro\ uses the
flux-limited diffusion approximation for gray or multigroup radiation.
The integration algorithm on the grid hierarchy is a recursive
procedure in which coarse grids are advanced in time, fine grids are
advanced multiple steps to reach the same time as the coarse grids and
the data at different levels are then synchronized. The
synchronization for self-gravity is similar to the algorithm
introduced by \cite{miniati-colella}.  Recent developments in
\castro\ include a deferred-correction method of coupling
hydrodynamics and reactions, a conservative gravity and rotation
source formulation~\cite{wdmergerI}, and retry mechanism to redo a
step based on criteria evaluated during the integration.\MarginPar{more new stuff?}

@@

\maestro\ and \castro\ share the same microphysics, available as the
\starkiller\ astrophysics github
project\footnote{\url{https://github.com/StarKiller-astro/microphysics/}}.
This includes equations of state and nuclear reaction
networks\footnote{Several of the reaction network righthand sides and
  the EOS were provided from Frank Timmes' {\em cococubed} software
  instruments page
  \url{http://cococubed.asu.edu/code_pages/codes.shtml}.  We thank him
  for making them available.}.  The reaction networks are written such
that that rates and integration strategy are decoupled, allowing us to
change the integration strategy for a set of rates.  They are also
written to be threadsafe and with GPUs in mind (more on that below).
The goal of \starkiller\ is to create a set of community microphysics
solvers that can be used in a variety of nuclear astrophysics codes,
not just those discussed here.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{sedov_scaling}
\includegraphics[width=0.48\linewidth]{wdmerger_scaling}
\caption{\label{fig:castro-scaling} (left) \castro\ strong scaling on
  OLCF titan for a pure hydro (Sedov w/ real EOS) problem.  The
  different colors represent different base resolutions.  We see
  excellent strong and weak scaling for the single level runs.  For
  the $1024^3$ run, we also ran with one $2\times$ jump level of
  refinment (triangles) and see good strong scaling.  The variability
  (shown by the error bars) at high core counts shows we are becoming
  work-starved. (right) \castro\ strong scaling on OLCF titan for a
  hydro + Poisson gravity (wdmerger) problem.  This problem uses a
  multipole solver to determine Dirichlet boundary conditions
  representing an isolated mass distribution, and then geometric
  multigrid to solve for the potential in the interior.  Two coarse
  grid sizes are shown, and demonstrate great strong scaling.  We also
  look at a single level of refinement (a $4\times$ jump) on top of
  this coarse grid.}
\end{figure}

@@

Figure~\ref{fig:maestro_scaling} shows strong scaling for \maestro\ on
the 3-d XRB problem.  This is a typical
\maestro\ application~\cite{xrb3}, where burning can be a significant
part of the overall evolution.  We ran on both OLCF titan and NERSC
edison.  We see that the code scales well to $\mathcal{O}(10^4)$
processors.  The upturn at the end of the scaling curves on titan
reflect the change from 1 MPI task / 8 OpenMP threads per NUMA node to
1 MPI task / 16 OpenMP threads per compute node (2 NUMA nodes).  The
main limitation to the scaling at the moment are the multigrid solves
used to enforce the projection (in particular the nodal solver).
\maestro\ also does not yet take advantage of the tiling capabilities
offered in \amrex.  These will be addressed in the future as we port
\maestro\ to the \cpp\ \amrex\ codebase to take advantage of ongoing
development.

Figure~\ref{fig:castro-scaling} shows scaling behavior for a pure
hydro + real EOS problem and a hydro / real EOS, + self-gravity
problem.  All tests were run on OLCF titan.  The AMD processors in
titan can be used in an 8 or 16 cores / node configuration.  For all
these runs, we ran in the 16 cores / node mode.  All runs also used
the PGI 17.7 compilers.  We see excellent scaling behavior for the
pure hydro problem.  For the self-gravity problem, we scale well with
a uniform grid, but with a $4\times$ jump for the AMR level on top of
this, we do worse at higher core counts.  This is a challenging
problem, as only 3\% of the domain volume is refined.  For a given
core count, there are a number of different combinations of MPI ranks
and OpenMP threads we can use.  In general, with multilevel problems,
we found the best performance with fewer MPI ranks and more OpenMP
threads.

Our latest focus has been on GPU ports of our application codes.  A
small proxy app, \starlord, was created from \castro\ with just the
hydrodynamics and stellar equation of state.  It uses a simple
method-of-lines formulation of the hydrodynamics and advects 13
nuclear species in addition to the hydrodynamics.  To offload work on
GPUs, GPU support was added directly into \amrex\footnote{This support
  is currently on a feature branch in the git repo, awaiting merge
  into {\tt development}.}.  In \amrex, grids consist of a number of
boxes that enable domain decomposition for parallelism.  An iterator
loops over all of the boxes at the same level of refinement and passes
a data pointer into a Fortran kernel function where the work is
done. For the simulation state data that resides in each box, we have
modified the memory allocator so that it can use a CUDA allocator
(either device or managed memory).  The domain iterator is configured
to handle data motion to and from the device, so that the compute
kernels can operate on data which is presumed to already be there, and
the computation is decoupled from the memory management. Computation
on the data can then be performed with OpenACC, CUDA Fortran, or (more
recently) OpenMP 4.5. We have also built CUDA compute support into
\amrex\ so that a compute kernel can be transparently operated on
using CUDA Fortran without substantially modifying the kernels (and we
anticipate using a similar strategy to use OpenACC and/or OpenMP in
the future). This helps ensure that we can continue to maintain
performance portability in our simulation codes, by decoupling the
physics algorithms from the backend support used to implement them on
various compute architectures.\MarginPar{add CPU time for reference}

@@

A parallel effort is porting our microphysics, in particular reaction
networks to GPUs.  Our strategy is to do the entire ODE integration on
the GPU, i.e., the data for a patch of zones is passed to the GPU, all
righthand size and Jacobian evaluations and the timestepping itself is
done on the GPU and once all zones are burned, we access the data as
needed on the CPU.  To enable this, we ported our workhorse ODE
integrator (VODE~\cite{vode}) to CUDA Fortran. This required extensive
rewrites of the internals of VODE, which was originally written using
Fortran 77 syntax unsupported by CUDA Fortran. Accelerating VODE with
CUDA Fortran has proven successful, and we now see significant
performance gains with the CUDA version of our reaction networks, even
for moderate sized networks (a 13-isotope standard network).
Figure~\ref{fig:cudaode} shows the speed-ups on a GPU vs.\ single CPU
core.

@@

We described our suite of astrophysics codes built on the
\amrex\ block-structured adaptive mesh refinement framework.  These
codes were developed to model problems in stellar astrophysics
spanning from low speed convection to explosive burning.  A major
theme of the codes is the open development model, with the code
development done on github and all problem files needed to recreate
any science results freely available.  Future development efforts for
\maestro\ include higher-order hydrodynamics and time-integration and
rotation.  For \castro, we are investigating stronger coupling between
hydrodynamics and reactions, new solvers (including MHD), and
finishing the GPU port.

\ack The work at Stony Brook was supported by DOE/Office of Nuclear
Physics grant DE-FG02-87ER40317 and NSF award AST-1211563.  An award
of computer time was provided by the Innovative and Novel
Computational Impact on Theory and Experiment (INCITE) program.  This
research used resources of the Oak Ridge Leadership Computing Facility
at the Oak Ridge National Laboratory, which is supported by the Office
of Science of the U.S. Department of Energy under Contract
No.\ DE-AC05-00OR22725.  We appreciate the efforts of the OLCF (and in
particular Fernanda Foertter) in organizing the GPU hackathons.  This
research used resources of the National Energy Research Scientific
Computing Center, which is supported by the Office of Science of the
U.S. Department of Energy under Contract No.\ DE-AC02-05CH11231.
Results in this paper were obtained using the high-performance LIred
computing system at the Institute for Advanced Computational Science
at Stony Brook University, which was obtained through the Empire State
Development grant NYS \#28451.  The GPU work benefited from a grant of
an Titan X Pascal board from Nvidia Corp through the GPU Grant
Program.  This research has made use of NASA's Astrophysics Data
System Bibliographic Services.

>>>>

The largest open theoretical question for SNe Ia is---what is the
nature of the progenitor?  About 20 years ago, the community had
mostly converged up the single-degenerate scenario---a
Chandrasekhar-mass C/O white dwarf that accretes from its companion,
eventually leading to runaway at the center that burns through the
star (see~\cite{hillebrandtniemeyer2000} for the state of the field at
that time).  Since then, the wealth of observations has show that
there is a lot of diversity in SNe Ia, and searches for progenitor
systems have strongly suggested that Chandra-mass white dwarfs cannot
explain most SNe Ia.  Today, merging white dwarfs (the double
degenerate scenario) have perhaps become the most popular model.
Other progenitors, like He burning on the surface of a sub-Chandra
white dwarf have also seen interest in explaining some of the observed
diversity.  See~\cite{araa-maoz} for a review.

There are open questions in all of these scenarios that can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is, can a deflagration
transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
possible to create a detonation in the thin surface He layer?  For
double degenerates, does the burning take place promptly? and can we
avoid an accretion-induced collapse to a neutron star.  For many of
these, we also need to address numerical issues, such as whether it is
possible to accurately model the ignition of a detonation with the
spatial resolution we can attain?  These are some of the questions we
seek to answer.

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model has \amrex\ managing the grid data-structures
and parallel communications, and it calls the computational kernels on
a patch-by-patch basis.  The core library is written in \cpp\ with 
computational kernels written in Fortran\footnote{Currently
  \maestro\ is using an older pure-Fortran interface, but this will be
  ported to the \cpp\ \amrex\ this coming year.}---this allows us to take
advantage of the strengths of both languages. \amrex\ supports subcycling
in time, which we use in \castro.

\amrex\ uses a hybrid MPI + OpenMP approach to parallelism.
Distribution of grid patches to nodes using MPI provides a natural
coarse-grained approach to distributing the computational work, while
subdividing patches into tiles and using threads to parallelize over
tiles using OpenMP provides effective fine-scale parallelization and
amortizes thread overhead over large units of work.  For \castro\,
this tiling logically subdivides AMR patches to increase the
parallelism for OpenMP, without incurring a large increase in metadata
that would be associated with using smaller patches in a flat MPI
mode~\cite{tiling}.  This strategy is especially important for many
core architectures like the Intel Phi.  Ongoing development is being
done to support GPU offloading, using managed memory provided by the
latest generations of GPUs.

There are many application codes built on \amrex, including those in
combustion, multiphase flow, accelerator design, and microfluidics. In
astrophysics, these include \maestro\ and \castro\ for stellar and
nuclear astrophysics applications and \nyx~\cite{nyx} for cosmological
applications.  Here we focus on the former two.

@@

The state is advanced using a second-order accurate projection method.
Fluid quantities are advected using an unsplit Godunov method, with
reactions incorporated via operator splitting.  The provisional
velocities are then projected onto the space that satisfies the
divergence constraint.  The projections require solving a variable
coefficient elliptic equations, which is done numerically using a
multigrid algorithm.  A number of recent advances in low Mach
number modeling~\cite{kleinpauluis,vasil:2013} have been
incorporated into \maestro.

@@

\castro~\cite{castro,castroII,castroIII} is a fully-compressible
radiation hydrodynamics code that supports arbitrary equations of
state, nuclear reaction networks, and Poisson gravity using geometric
multigrid.  The main hydrodynamics scheme in \castro\ is an unsplit
piecewise parabolic method.  The radiation solver in \castro\ uses the
flux-limited diffusion approximation for gray or multigroup radiation.
The integration algorithm on the grid hierarchy is a recursive
procedure in which coarse grids are advanced in time, fine grids are
advanced multiple steps to reach the same time as the coarse grids and
the data at different levels are then synchronized. The
synchronization for self-gravity is similar to the algorithm
introduced by \cite{miniati-colella}.  Recent developments in
\castro\ include a spectral-deferred correction method of coupling
hydrodynamics and reactions, a conservative gravity and rotation
source formulation~\cite{wdmergerI}, and retry mechanism to redo a
step based on criteria evaluated during the integration.\MarginPar{more new stuff?}

@@

\maestro\ and \castro\ share the same microphysics, available as the
\starkiller\ astrophysics github
project\footnote{\url{https://github.com/StarKiller-astro/microphysics/}}.
This includes equations of state and nuclear reaction
networks\footnote{Several of the reaction network righthand sides and
  the EOS were provided from Frank Timmes' {\em cococubed} software
  instruments page
  \url{http://cococubed.asu.edu/code_pages/codes.shtml}.  We thank him
  for making them available.}.  The reaction networks are written such
that the rates and integration strategy are decoupled, allowing us to
change the integration strategy for a given set of rates.  They are
also written to be threadsafe and with GPUs in mind (more on that
below).  The goal of \starkiller\ is to create a set of community
microphysics solvers that can be used in a variety of nuclear
astrophysics codes, not just those discussed here.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{sedov_scaling}
\includegraphics[width=0.48\linewidth]{wdmerger_scaling}
\caption{\label{fig:castro-scaling} (left) \castro\ strong scaling on
  OLCF titan for a pure hydro (Sedov w/ real EOS) problem.  The
  different colors represent different base resolutions.  We see
  excellent strong and weak scaling for the single level runs.  For
  the $1024^3$ run, we also ran with one level of refinment by a
  factor of 2 (triangles) or a factor of 4 (squares), and see good
  strong scaling.  The variability (shown by the error bars) at high
  core counts shows we are becoming work-starved. (right)
  \castro\ strong scaling on OLCF titan for a hydro + Poisson gravity
  (wdmerger) problem.  This problem uses a multipole solver to
  determine Dirichlet boundary conditions representing an isolated
  mass distribution, and then geometric multigrid to solve for the
  potential in the interior.  Two coarse grid sizes are shown, and
  demonstrate great strong scaling.  We also look at a single level of
  refinement (by a factor of 4) on top of this coarse grid.}
\end{figure}

@@

Figure~\ref{fig:maestro_scaling} shows strong scaling for \maestro\ on
the 3-d XRB problem.  This is a typical
\maestro\ application~\cite{xrb3}, where burning can be a significant
part of the overall evolution.  We ran on both OLCF titan and NERSC
edison.  We see that the code scales well to $\mathcal{O}(10^4)$
processors.  The upturn at the end of the scaling curves on titan
reflect the change from 1 MPI task / 8 OpenMP threads per NUMA node to
1 MPI task / 16 OpenMP threads per compute node (2 NUMA nodes).  The
main limitation to the scaling at the moment is the multigrid solves
used to enforce the projection (in particular the nodal solver).
\maestro\ also does not yet take advantage of the tiling capabilities
offered in \amrex.  These will be addressed in the future as we port
\maestro\ to the \cpp\ \amrex\ codebase to take advantage of ongoing
development.

Figure~\ref{fig:castro-scaling} shows scaling behavior for a pure
hydro + real EOS problem and a hydro / real EOS, + self-gravity
problem.  All tests were run on OLCF titan.  The AMD processors in
titan can be used in an 8 or 16 cores / node configuration.  For all
these runs, we ran in the 16 cores / node mode.  All runs also used
the PGI 17.7 compilers.  We see excellent scaling behavior for the
pure hydro problem.  For the self-gravity problem, we scale well with
a uniform grid, but with a factor of 4 refinement for the AMR level on
top of this, we observe degraded performance at higher core counts.
This is a challenging problem, as only 3\% of the domain volume is
refined.  For a given core count, there are a number of different
combinations of MPI ranks and OpenMP threads we can use.  In general,
with multilevel problems, we found the best performance with fewer MPI
ranks and more OpenMP threads.

Our latest focus has been on GPU ports of our application codes.  A
small proxy app, \starlord, was created from \castro\ with just the
hydrodynamics and stellar equation of state.  It uses a simple
method-of-lines formulation of the hydrodynamics and advects 13
nuclear species in addition to the hydrodynamics.  To offload work on
GPUs, GPU support was added directly into \amrex\footnote{This support
  is currently on a feature branch in the git repo, awaiting merge
  into {\tt development}.}.  In \amrex, grids consist of a number of
boxes that enable domain decomposition for parallelism.  An iterator
loops over all of the boxes at the same level of refinement and passes
a data pointer into a Fortran kernel function where the work is
done. For the simulation state data that resides in each box, we have
modified the memory allocator so that it can use a CUDA allocator
(either device or managed memory).  The domain iterator is configured
to handle data motion to and from the device, so that the compute
kernels can operate on data that is presumed to already be there, and
the computation is decoupled from the memory management. Computation
on the data can then be performed with OpenACC, CUDA Fortran, or (more
recently) OpenMP 4.5. We have also built CUDA compute support into
\amrex\ so that a compute kernel can be transparently operated on
using CUDA Fortran without substantially modifying the kernels (and we
anticipate using a similar strategy to use OpenACC and/or OpenMP in
the future). This helps ensure that we can continue to maintain
performance portability in our simulation codes, by decoupling the
physics algorithms from the backend support used to implement them on
various compute architectures.\MarginPar{add CPU time for reference}

@@

A parallel effort is porting our microphysics, in particular reaction
networks to GPUs.  Our strategy is to do the entire ODE integration on
the GPU, i.e., the data for a patch of zones is passed to the GPU, all
righthand size and Jacobian evaluations and the timestepping itself is
done on the GPU and once the burning in all zones is completed, we access the data as
needed on the CPU.  To enable this, we ported our workhorse ODE
integrator (VODE~\cite{vode}) to CUDA Fortran. This required extensive
rewrites of the internals of VODE, which was originally written using
Fortran 77 syntax unsupported by CUDA Fortran. Accelerating VODE with
CUDA Fortran has proven successful, and we now see significant
performance gains with the CUDA version of our reaction networks, even
for moderate sized networks (a 13-isotope standard network).
Figure~\ref{fig:cudaode} shows the speed-ups on a GPU vs.\ single CPU
core.

@@

We have described our suite of astrophysics codes built on the
\amrex\ block-structured adaptive mesh refinement framework.  These
codes were developed to model problems in stellar astrophysics
spanning from low speed convection to explosive burning.  A major
theme of the codes is the open development model, with the code
development done on github and all problem files needed to recreate
any science results freely available.  Future development efforts for
\maestro\ include higher-order hydrodynamics and time-integration and
rotation.  For \castro, we are investigating stronger coupling between
hydrodynamics and reactions, new solvers (including MHD), and
finishing the GPU port.

\ack The work at Stony Brook was supported by DOE/Office of Nuclear
Physics grant DE-FG02-87ER40317 and NSF award AST-1211563.  The work
at LBNL was supported by the DOE Office of Advanced Scientific
Computing Research under Contract No, DE-AC02-05CH11231. An award of
computer time was provided by the Innovative and Novel Computational
Impact on Theory and Experiment (INCITE) program.  This research used
resources of the Oak Ridge Leadership Computing Facility at the Oak
Ridge National Laboratory, which is supported by the Office of Science
of the U.S. Department of Energy under Contract
No.\ DE-AC05-00OR22725.  We appreciate the efforts of the OLCF (and in
particular Fernanda Foertter) in organizing the GPU hackathons.  This
research used resources of the National Energy Research Scientific
Computing Center, which is supported by the Office of Science of the
U.S. Department of Energy under Contract No.\ DE-AC02-05CH11231.
Results in this paper were obtained using the high-performance LIred
computing system at the Institute for Advanced Computational Science
at Stony Brook University, which was obtained through the Empire State
Development grant NYS \#28451.  The GPU work benefited from a grant of
an Titan X Pascal board from Nvidia Corp through the GPU Grant
Program.  This research has made use of NASA's Astrophysics Data
System Bibliographic Services.
